Citation Analysis of Computer Science Publications

### 2019-12-13 Scientometrics/comp.tex
update
updated script, reordered figures

Figure~\ref{fig:heatmap} permits examination of these data from two perspectives: (i) ASJC minor subject areas that comprised at least 15\% of the publication in a cluster (ii) the clusters that map to a given ASJC minor subject area. Under these conditions, 31 of the 334 ASJC minor subject areas are represented. Unsurprisingly, the broad category Computer Science Applications was present in 16 of 20 clusters, followed by Software in 12 clusters, and Electrical and Electronic Engineering in 10 clusters.

From the other perspective, cluster 17 was the most diverse and contained publications annotated with 8 minor subject area labels, of which Biochemistry, Chemistry(all), Genetics, and Molecular Biology were the most distant from computer science while Statistics \& Probability, Computational Theory \& Mathematics, and Computational Mathematics are closer.  These data suggest that areas central to computer science in 2019 (Salton and Bergmark's historical triad of hardware, software, and theory) are more likely to be found in multiple clusters, whereas, biology, chemistry, and statistics are more restricted in distribution.  A second inference is that, in a few cases journal-based classification and article clusters are tightly related in terms of article content (Hardware and Architecture). Thirdly, that some areas of computer science are relatively ubiquitous, e.g. computer applications. Last, that interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters, e.g., cluster 17 (ibid) and cluster 18 (artificial intelligence, cognitive neuroscience, and neurology).

We also attempted to map these 20 clusters assembled through direct citation to the CCS classification by manually matching the top 25-50 most heavily cited publications in each cluster to corresponding categories in the CCS. This was feasible in some cases, clusters 0 and 1 mapped reasonably well to the top level categories Hardware and Computer Systems Organization but less so in the cases of other clusters. Further, the top cited papers often derived from biology and we were concerned that this small sample (top cited papers) were not representative of the cluster. Similar challenges inherent in labeling citation clusters at a high level of aggregation have also been noted~\cite{waltman_new_2012}.

\subsection{Clustering by Co-citation.} Co-citation was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, and provides insight into the emergence of new ideas derived from the association of previously independent ones.  Co-citation thus provides some degree of evolutionary insight and has been used for clustering publications~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}. As described in Methods, we performed 600 iterations of agglomerative clustering to the 10,298 clusters. To focus on major clusters, we restricted agglomeration to clusters containing at least 10 nodes, which resulted in 666 clusters. Further agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}).  Consequently, the 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136.

@@

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
 \includegraphics[scale=0.5]{cocit_venn2.pdf}
% figure caption is below the figure
\caption{High-level disciplinary composition of publications in $\sim$2,000 clusters generated by variable-level clustering  of co-citation data.
%Variable Level Clustering. 
84,591 nodes from were matched to the 4 subject areas, 
27 major subject areas, and 334 minor subject areas in the Scopus ASJC classification. Since 82.5\% of these nodes were assigned more to than one minor subject category, 
we assigned fractional counts for the four subject areas that summed to 1 for each article. The figure shows a four way Venn Diagram~\cite{chen_venn_2011} illustrating the 
relative frequencies of publications from the Physical Sciences, Social Sciences, Life Sciences, and Health Sciences.}
\label{fig:cocitr_venn2}       % Give a unique label
\end{figure}

>>>>

Figure~\ref{fig:heatmap} permits examination of these data from two perspectives: (i) rows: the clusters that map to a given ASJC minor subject area (ii) columns: ASJC minor subject areas that comprised at least 15\% of the publication in a cluster. Under these conditions of clustering and a threshold of 15\%, 31 of the 334 ASJC minor subject areas are represented. Unsurprisingly, the broad categories Computer Science Applications, Software, and Electrical and Electronic Engineering are seen in 17, 12, and 10 clusters respectively while Artificial Intelligence mapped to 7 different clusters. At the other end of the range, 12 of the 31 ASJC minor subject areas were each detected in a single cluster.

From the alternate perspective (columns), cluster 17 was the most diverse and contained publications annotated with 8 minor subject area labels: Biochemistry, Chemistry(all), Genetics, Molecular Biology,  Statistics \& Probability, Computational Theory \& Mathematics, and Computational Mathematics. Cluster 19 mapped to two areas but was excluded from qualitative analysis because of its high conductance value. Of the remaining clusters, the third (Cluster 2) represents interactions between the minor subject areas Theoretical Computer Science, Discrete Mathematics and Combinatorics, and Applied Mathematics as well as the more generic Computer Science (all). Clusters 3 and 4 include Computer Networks and Communications and Cluster 18 (Artificial intelligence, Cognitive neuroscience, and Neurology) and clusters 5-9 include Management Science, Operations Research, Information Systems, Modeling and Simulation, and Human Computer Interaction.

To us, these data suggest that areas central to computer science in 2019 (Salton and Bergmark's historical triad of hardware, software, and theory) are more likely to be found in multiple clusters, whereas areas such as biology, chemistry, and statistics are more restricted in distribution.  A second inference is that, in some cases journal-based classification and our article clusters are similar in terms of article content (Hardware and Architecture). Thirdly, that some areas of computer science are relatively ubiquitous, e.g. computer applications. Lastly, that interactions with fields outside computer science are restricted to a single cluster or small numbers of clusters, e.g., cluster 17  and cluster 18.

We also attempted to map these 20 clusters assembled through direct citation to the CCS classification by manually matching the top 25-50 most heavily cited publications in each cluster to corresponding categories in the CCS. This was feasible in some cases, clusters 0 and 1 mapped reasonably well to the top level categories Hardware and Computer Systems Organization but much less so in the cases of other clusters. Further, the top cited papers often derived from biology and we were concerned that they were not representative of the cluster. The challenges inherent in labeling citation clusters at a high level of aggregation have been previously noted~\cite{waltman_new_2012}.

[closing sentence necessary?]

\subsection{Clustering by Co-citation.} Co-citation was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, and provides insight into the emergence of new ideas derived from the association of previously independent ones.  Co-citation thus provides some degree of evolutionary insight and has been used for clustering publications~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}. Unlike, clustering by direct citation using Graclus, which preserves all publications in \emph{comp}, the process of clustering by co-citation is progressively selective. As described in Methods, we first identified the top 10\% most-cited articles in the `\emph{comp} dataset and then documented all co-cited pairs of their cited references that were also elements of \emph{comp}. The frequency of co-citation was then computed across Scopus \emph{comp} and normalized using Salton's correction. These data were then represented as a graph where each node was a publication and each edge, the normalized co-citation frequency of a co-cited pair.  To cluster, we performed variable level clustering (VLC) followed by agglomerative clustering of the clusters resultant from VLC. To focus on major clusters, we also restricted agglomeration to clusters containing at least 10 nodes, which resulted in 666 clusters. Further agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}).  Consequently, the 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136, with total nodes summing to 8,042.

@@

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
 \includegraphics[scale=0.5]{cocit_venn2.pdf}
% figure caption is below the figure
\caption{High-level disciplinary composition of publications in $\sim$2,000 clusters generated by variable-level clustering  of co-citation data.
%Variable Level Clustering. 
84,591 nodes from were matched to the 4 subject areas, 
27 major subject areas, and 334 minor subject areas in the Scopus ASJC classification. Since 82.5\% of these nodes were assigned more to than one minor subject category, 
we assigned fractional counts for the four subject areas that summed to 1 for each article. The figure shows a four way Venn Diagram~\cite{chen_venn_2011} illustrating the 
relative frequencies of publications from the Physical Sciences, Social Sciences, Life Sciences, and Health Sciences.}
\label{fig:cocitr_venn2}       % Give a unique label
\end{figure}
\newpage
